---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "65%"
)
```

# grobblR

<!-- badges: start -->
<!-- badges: end -->

The goal of grobblR is to give R users the ability to easily create flexible, reproducible PDF reports incorporating *grobs* and allowing the user to combine them in any way they desired. *Grobs* are taken from `grid::grob()` [documentation](https://stat.ethz.ch/R-manual/R-devel/library/grid/html/grid.grob.html):

> A grid graphical object (“grob”) is a description of a graphical item. These basic classes provide default behaviour for validating, drawing, and modifying graphical objects. 

Within grobblR, the objects able to be converted to a grob are:  

  * Data frames / matrices
      + A vector of values
  * `ggplot` plots  
  * `.png` files  
  * Character strings  
  * Empty space (represented with `NA`)

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("calvinmfloyd/grobblR")
```

***

## Grob Layout

* We decided on a Shiny-like bootstrap method to arrange grobs, which would incorporate nested grob-rows (a row combining one or several grobs) and grob-columns (a column combining one or several grobs) within an overall layout of grobs (which we will call a grob-layout).
* The main grob-layout function evaluates the overall dimensions of the page you desire, and then distributes heights and widths to the grobs that will make up the page, based on proportions and the amount of grob-rows (`grob_row()`) and grob-columns (`grob_col()`) within the overall grob-layout (`grob_layout()`).
* For example, if we wanted to do a simple `2 x 2` grid system on a `100mm x 100mm` (millimeters is the default unit in grobblR) page, we would simply write:

```{r 2x2 system}

library(grobblR)
library(gridExtra)
 
gl = grob_layout(
  grob_row(grob_col(1), grob_col(2)),
  grob_row(grob_col(3), grob_col(4)),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

* The `grob_row()`'s tell the `grob_layout()` that we want two rows on the outermost layer, and since there are two `grob_col()`'s within each of the `grob_row()`'s, the result is a `2 x 2` grid.
* But the result doesn't have to be a nice `2 x 2` grid - we could have a layout where the first row has two columns and the second has one:

```{r 2 row, 3 element system}
gl = grob_layout(
  grob_row(grob_col(1), grob_col(2)),
  grob_row(grob_col(3)),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## Nested Rows and Columns

* One of the ways that makes grobblR truly flexible is the ability to create "nested" rows and columns - where we can insert grob-rows into grob-columns, and vice-versa.
* This is made possible by incorporating the active classes within the [R6 package](https://cran.r-project.org/web/packages/R6/index.html).
* The above grob with a second grob-column in the second grob-row which contains two grob-rows within it:

```{r 2x2 system w/ borders}
gl = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(border = TRUE, 1),
    grob_col(border = TRUE, 2)
    ),
  grob_row(
    border = TRUE,
    grob_col(border = TRUE, 3),
    grob_col(
      border = TRUE,
      grob_row(border = TRUE, grob_col(border = TRUE, 4)),
      grob_row(border = TRUE, grob_col(border = TRUE, 5))
      )
    ),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

### Size Control

* One way to control heights and widths of the grobs is by tweaking the `p` (standing for *proportion*) parameter within both `grob_row()` and `grob_col()`.
* The default for `p` is 1, but let's see what happens when we tweak it for a grid-system with two grob-rows:

```{r 2x1 system - p parameter}
gl = grob_layout(
  grob_row(p = 1, border = TRUE, grob_col('1')),
  grob_row(p = 2, border = TRUE, grob_col('2')),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

* The user can also specify specific heights/widths in the unit the user chooses (either "mm", "cm" or "inches").

```{r 3x1 system - specific heights}
gl = grob_layout(
  grob_row(height = 25, border = TRUE, grob_col('1')),
  grob_row(height = 50, border = TRUE, grob_col('2')),
  grob_row(height = 25, border = TRUE, grob_col('3')),
  height = 100,
  width = 100,
  padding = 0
  )

grid.arrange(gl$grob)
```

* Or the user can use a combination of specific heights and proportions to size the grobs:

```{r 3x1 system - combination}
gl = grob_layout(
  grob_row(p = 3, border = TRUE, grob_col('1')),
  grob_row(height = 50, border = TRUE, grob_col('2')),
  grob_row(p = 1, border = TRUE, grob_col('3')),
  height = 100,
  width = 100,
  padding = 0
  )

grid.arrange(gl$grob)
```

## Matrix Grobs

* Using the `aes_list` parameter and the `ga_list()` function within `grob_col()`, the aesthetics of individual grobs can be adjusted to how the user intends them to appear.
* See `?grobblR::ga_list` for a full list and description for each of the possible aesthetic options.
* For uniform background color for each cell in a matrix, all the user would need to do is set a single value for `background_color` as an element within `aes_list`:


```{r First matrix example}
mat = matrix(1:4, nrow = 2, byrow = TRUE)
gl = grob_layout(
  grob_row(grob_col(mat, aes_list = ga_list(background_color = 'gray90'))),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

* If the user wants different cells of the matrix to have different background colors, there needs to be a separate aesthetic matrix corresponding to the background color of each cell of the original matrix
* Since `mat` is a `2 x 2` matrix, we need to specify a another `2 x 2` matrix with each element denoting a background color for each of the cells in `mat`, incorporating the `aes_matrix()` `alter_columns()` and `alter_cells()` functions:

```{r First matrix, cell-by-cell customization, 1}
`%>%` = dplyr::`%>%`
color_mat = mat %>% aes_matrix('white') %>% alter_columns('red', 1) %>% alter_cells('gray90', 1, 2)

gl = grob_layout(
  grob_row(grob_col(mat, aes_list = ga_list(background_color = color_mat))),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## ggplot Grobs

* The `ggplot` plot is stretched or squished depending on what the dimensions of the allotted space are:

```{r ggplot Grobs}
data(iris)
library(ggplot2)
gg1 = ggplot(
  data = iris, 
  mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)
  ) +
  geom_point() +
  guides(color = FALSE)

gg2 = ggplot(
  data = iris,
  mapping = aes(x = Sepal.Length, y = Petal.Length, color = Species)
  ) +
  geom_point() +
  guides(color = FALSE)

gl = grob_layout(
  grob_row(grob_col(gg1), grob_col(gg2)),
  grob_row(grob_col(head(iris))),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## Image Grobs

* Only images in `.png` format are allowed in grobblR. 
* If the user does not want to maintain the aspect ratio of the image and allow it to be stretched within its allotted space, `maintain_aspect_ratio = F` must be inserted within the `aes_list` list.

```{r .png Grobs}
gl = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(
      border = TRUE,
      'vignettes/kings_logo.png'
      ),
    grob_col(
      border = TRUE,
      aes_list = ga_list(maintain_aspect_ratio = FALSE),
      'vignettes/kings_logo.png'
      )
    ),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## Character String Grobs

* grobblR handles long character strings by making the text as big as possible, while still being able to fit the text inside its allotted space.
* This results in automatically putting different words on different lines:

```{r simple text grob}
text = "The quick brown fox jumps over the lazy dog."

gl = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(
      border = TRUE,
      text
      )
    ),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## Empty Space Grobs

* Empty space grobs in case the user needs to compact, or shift grobs around:

```{r empty space grob}

df = data.frame(letter = letters[1:5], col1 = 1:5, col2 = 5:1)

gl = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(df),
    grob_col(
      grob_row(grob_col(df)),
      grob_row(grob_col(p = 1/3, NA))
      ),
    grob_col(
      grob_row(grob_col(p = 1/3, NA)),
      grob_row(grob_col(df))
      )
    ),
  height = 100,
  width = 100
  )

grid.arrange(gl$grob)
```

## Grob Layout to PDF

* The default values for the `height` and `width` parameters within `grob_layout()` are `280` and `216` and the default units are `mm`, as these are the values needed to properly fit the grob-layout onto a piece of standard computer paper, portrait orientation. 
* Converting multiple grob-layout's into a single PDF report using `grob_to_pdf()`, with a file title and a meta data title:

```{r Converting grob-layout to PDF, eval=FALSE}

first_page_grob_layout = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(df),
    grob_col(
      grob_row(grob_col(df)),
      grob_row(grob_col(p = 1/3, NA))
      ),
    grob_col(
      grob_row(grob_col(p = 1/3, NA)),
      grob_row(grob_col(df))
      )
    ),
  height = 100,
  width = 100
  )

second_page_grob_layout = grob_layout(
  grob_row(
    border = TRUE,
    grob_col(
      border = TRUE,
      text
      )
    ),
  height = 100,
  width = 100
  )

# grob_to_pdf(
#   first_page_grob_layout,
#   second_page_grob_layout,
#   file_name = 'test.pdf',
#   meta_data_title = 'Test PDF'
#   )

# OR

grob_to_pdf(
  list(first_page_grob_layout, second_page_grob_layout),
  file_name = 'test.pdf',
  meta_data_title = 'Test PDF'
  )

```






