---
title: "Grob Layout Customization"
author: "Calvin Floyd"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grob Layout Customization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4
)
```


* Arguably the most crucial function in `grobblR` is `grob_layout()`, which needs to be utilized in order to create the final PDF.
* `grob_layout()` reads the grob-row's and grob-columns within it to know how to _layout_ the page. See the [grobblR vignette](grobblR.html) for information on how it allots space to the grob-rows and grob-columns.
* There are many ways customize individual grob-rows and grob-columns within the grob-layout, including borders, titles and captions.

```{r load}
library(grobblR)
```

## Borders

* Within `grob_layout()`, the `grob_row()` and `grob_col()` functions are able to put borders around their contents.

```{r two_by_two_system_w_borders}
grob_layout(
  grob_row(
    grob_col(1),
    grob_col(border = TRUE, 2)
    ),
  grob_row(
    border = TRUE,
    grob_col(3),
    grob_col(
      border = TRUE,
      grob_row(border = TRUE, grob_col(border = TRUE, 4)),
      grob_row(border = TRUE, grob_col(border = TRUE, 5))
      )
    ),
  height = 100,
  width = 100
  ) %>%
  view_grob()
```

* Those borders are customizable. With the `border_aes_list` argument and the `ga_list()` function, the user can adjust
  * Which border sides to display (`border_sides`)
  * The border width (`border_width`)
  * The border color (`border_color`)
* For more information on the aesthetics above, see `?ga_list`.

```{r customizable_borders}
grob_layout(
  grob_row(
    grob_col(1),
    grob_col(
      border = TRUE,
      border_aes_list = ga_list(
        border_color = "red"
        ),
      2
      )
    ),
  grob_row(
    border = TRUE,
    border_aes_list = ga_list(
      border_sides = "top",
      border_width = 5
      ),
    grob_col(3),
    grob_col(
      border = TRUE,
      border_aes_list = ga_list(
        border_sides = "left, bottom",
        border_color = "blue",
        border_width = 4
        ),
      4
      )
    ),
  height = 100,
  width = 100
  ) %>%
  view_grob()
```

##  Titles

* Titles can be added to either `grob_layout()`, `grob_row()` or `grob_col()`.

```{r title}
grob_layout(
  title = "grob-layout title",
  grob_row(
    grob_col(1),
    grob_col(title = "grob-column title", 2)
    ),
  grob_row(
    title = "grob-row title",
    grob_col(3),
    grob_col(4)
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

* The height of the title can be adjusted by either giving an explicit height in millimeters to the `title_height` argument, or changing the title height proportion with `title_p`.
  * `title_p` corresponds to the proportion of the height allotted to the grob in question which will be given to the title.
  *  `title_height` will override `title_p` in all cases.
  
```{r title_height_adjustment}
grob_layout(
  title = "grob-layout title",
  title_height = 20,
  grob_row(
    grob_col(1),
    grob_col(
      title = "grob-column title",
      title_p = 0.5,
      2
      )
    ),
  grob_row(
    title = "grob-row title",
    title_height = 10,
    grob_col(3),
    grob_col(4)
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

* The aesthetics of the title can be altered as well with the `title_aes_list` argument and `ga_list()`.

```{r title_aesthetic_adjustment}
grob_layout(
  title = "grob-layout title",
  title_height = 20,
  title_aes_list = ga_list(
    text_color = "blue",
    border_sides = "bottom"
    ),
  grob_row(
    grob_col(1),
    grob_col(
      title = "grob-column title",
      title_p = 0.5,
      title_aes_list = ga_list(
        font_face = 3,
        text_color = "white",
        background_color = "gray40"
        ),
      2
      )
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

## Captions

* Everything that applies to titles also applies to captions, as `grob_layout()`, `grob_col()` and `grob_row()` are all equipped with `caption`, `caption_aes_list`,  `caption_p` and `caption_height` arguments.
* By default, captions are more subtle than titles, but their appearances can be altered however the user wants.

```{r caption}
grob_layout(
  caption = "grob-layout caption",
  grob_row(
    grob_col(1),
    grob_col(caption = "grob-column caption", 2)
    ),
  grob_row(
    caption = "grob-row caption",
    grob_col(3),
    grob_col(4)
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

  
```{r caption_height_adjustment}
grob_layout(
  caption = "grob-layout caption",
  caption_height = 20,
  grob_row(
    grob_col(1),
    grob_col(
      caption = "grob-column caption",
      caption_p = 0.5,
      2
      )
    ),
  grob_row(
    caption = "grob-row caption",
    caption_height = 10,
    grob_col(3),
    grob_col(4)
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

```{r caption_aesthetic_adjustment}
grob_layout(
  caption = "grob-layout caption",
  caption_height = 20,
  caption_aes_list = ga_list(
    text_color = "blue",
    border_sides = "bottom"
    ),
  grob_row(
    grob_col(1),
    grob_col(
      caption = "grob-column caption",
      caption_p = 0.5,
      caption_aes_list = ga_list(
        font_face = 3,
        text_color = "white",
        background_color = "gray40"
        ),
      2
      )
    )
  ) %>%
  view_grob(height = 100, width = 100)
```

## Page Numbers

* The user can add a page number to the bottom right of the page if he/she wishes, by utilizing the `page_number` argument within `grob_layout()`.
* The page number is inserted into the padding of the overall grob-layout.
* As of now, the aesthetics of the page number is restricted to the default behavior.

```{r page_numbers}
grob_layout(
  grob_row(
    grob_col(1),
    grob_col(2)
    ),
  grob_row(
    grob_col(3),
    grob_col(4)
    ),
  page_number = 1
  ) %>%
  view_grob(height = 100, width = 100)
```

* FYI - The page number allotted space is more meant for a fully-sized grob-layout (using default grob-layout size) than it is for a 100mm x 100mm sized grob.


